\documentclass[handout]{beamer} %
\usetheme{CambridgeUS}
\usepackage[latin1]{inputenc}
\usefonttheme{professionalfonts}
\usepackage{times}
\usepackage{graphics}
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{booktabs}

\usepackage[style=authoryear]{biblatex}

\addbibresource{references.bib}

% theorem
\BeforeBeginEnvironment{theorem}{
  \setbeamercolor{block title}{fg=red!80!black,bg=gray!10!white}
  \setbeamercolor{block body}{fg=black,bg=white}
}
\AfterEndEnvironment{theorem}{
    \setbeamercolor{block title}{fg=blocktitlefgsave,bg=blocktitlebgsave}
    \setbeamercolor{block body}{fg=blockbodyfgsave,bg=blockbodybgsave}
}


% definition
\BeforeBeginEnvironment{definition}{
  \setbeamercolor{block title}{fg=red!80!black,bg=gray!10!white}
  \setbeamercolor{block body}{fg=black, bg=white}
}
\AfterEndEnvironment{definition}{
    \setbeamercolor{block title}{fg=blocktitlefgsave,bg=blocktitlebgsave}
    \setbeamercolor{block body}{fg=blockbodyfgsave,bg=blockbodybgsave}
}

\renewcommand{\arraystretch}{1.5}

%\setbeamersize{text margin left=0.3mm,text margin right=0.1mm}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\expit}{expit}
\DeclareMathOperator*{\Cov}{Cov}
\DeclareMathOperator*{\Var}{Var}
\newcommand{\R}{\mathbb{R}}

\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=blue, filecolor=magenta, urlcolor=blue}

\author[CL]{Caleb Leedy}
\title[Research]{New Research in Causal Inference}

\begin{document}

\everymath{\displaystyle}
\setbeamertemplate{title page}[default][colsep=-4bp,rounded=true]
\setbeamertemplate{itemize items}[circle]
\setbeamercolor{block title}{bg=red!50,fg=black}
\frame{\titlepage}

\begin{frame}{Causal Inference Research Ideas}

\begin{itemize}
  \item<1-> Ideally, everyone is coming today with new ideas for research in causal
    inference.
  \item<2-> Even if you did not there is hope!
\end{itemize}

\end{frame}

\begin{frame}{Stages of Research Ideas}

  \begin{itemize}
    \item[1.] Topic of Interest (Ex. Causal inference with spatial data.)
    \item[2.] Problem of Interest (Ex. Spatial treatment effects.)
    \item[3.] Idea for a Solution (Ex. Use a spatial varying coefficient model.
      Challenge is probably defining the missing mechanism.)
    \item[4.] Initial Results
  \end{itemize}

\end{frame}

\begin{frame}{Types of Ideas}

  \begin{itemize}
    \item<1-> Not every idea is good.
    \item<2-> We love bad ideas.
  \end{itemize}

\end{frame}

\begin{frame}{Group Step 1}

  \begin{itemize}
    \item As a group of 2-3, share your research ideas together.
    \item Start with your bad ideas.
    \item Give constructive feedback to group member.
    \item Encourage them to be more detailed.
  \end{itemize}

\end{frame}

\begin{frame}{Group Step 2}

  \begin{itemize}
    \item As a large group, we will share our research ideas together.
    \item Give constructive feedback to group member.
    \item Encourage them to be more detailed.
  \end{itemize}

\end{frame}

%\begin{frame}{}
%
%  \begin{itemize}
%    \item
%  \end{itemize}
%
%\end{frame}

\end{document}
